Multi-wavelength Radio Continuum Emission Studies of Dust-free Red Giants

### 2013-04-05 draft_06/draft.tex
Edits on April 5th

\subsection{Radio Spectral Indices} \label{disc:disc3}
The radio emission from non-dusty K spectral-type red giants is due to thermal free-free emission in their partially ionized outflows. The radio flux density-frequency relationship for these type of stars is always found to be intermediate between that expected from an optically thick homogeneous plasma ($\alpha$ follows the Rayleigh-Jeans tail of the Planck function i.e., $\alpha = +2$) and that from an optically thin plasma ($\alpha = -0.1$).  It can be shown that the expected radio spectrum from an isothermal spherically symmetric steady flow varies as $\nu ^{0.6}$ \citep{1975MNRAS.170...41W,1975AA....39..217O,1975AA....39....1P}. If we relax some of these assumptions about the outflow and instead assume that the electron density and temperature vary as a function of distance from the star $r$, and have the form $n_{e} \propto r^{-p}$ and $T_{e} \propto r^{-n}$ respectively, then

@@

\begin{equation}
\alpha = \frac{4p -6.2 -0.6n}{2p-1-1.35n}
\label{eq:eq1}
\end{equation}
\citep[e.g.,][]{1987ApJ...312..813S}. For an isothermal flow $p > 1.5$, otherwise we are in the optically thin regime. Equation \ref{eq:eq1} is also only valid if the ionization fraction is constant throughout the outflow. By making these assumptions and assuming either a constant velocity (i.e., $p=2$) or an isothermal outflow (i.e., $n=0$), we can easily estimate the power-law decay of either the temperature or velocity profile, respectively.

@@

The higher degree of ionization in the mass outflow of $\alpha$ Boo in comparison to $\alpha$ Tau makes a substantial difference to the size of the stars effective radius as can be seen in Figure \ref{fig:fig4}. At 6, 13, and 20 cm the effective radius of $\alpha$ Boo at $\tau _{\rm{rad}}$=1/3 is predicted to be 1.6, 2.8, and 3.7 $R_{\star}$ but is only $\sim$1.1 $R_{\star}$ at 6 and 13 cm for $\alpha$ Tau. \cite{1998ApJ...503..396R} predict that $\alpha$ Tau's wind reaches $\sim$80\% of its terminal velocity by 3 $R_{\star}$ so it is highly unlikely that even our longest wavelength VLA measurements of $\alpha$ Tau sample its outer wind. For $\alpha$ Boo however, \cite{1985pssl.proc..351D} predicts that the wind has reached its terminal velocity by $\sim$2 $R_{\star}$ so based on this model our longest VLA measurements probable sample the wind at, or very close to, its terminal velocity. Assuming this to be the case, Equation \ref{eq:eq1} then indicates that $\alpha$ Boo's steady outflow undergoes rapid cooling at a few stellar radii. The derived temperature coefficient of 1.65 points to a temperature power-law falloff exceeding that from a wind adiabatically cooling with no heat source (i.e. n=4/3 in this case) with the excess cooling being possible due to recombination of H$^{+}$ or heavier ions. Finally, if the wind ionization balance has not yet become \textit{frozen-in} in the region of $\alpah$ Boo's wind where the radio emission emanates from then the excess slope of the spectral index could be due to a combination of both cooling and changing ionization balance. In this scenario the temperature coefficient may be smaller then our derived value here.

@@

A failure of the existing atmospheric model for $\alpha$ Boo is that is overestimates the radio fluxes at long VLA wavelengths which sample its outer atmosphere, as clearly shown in Figure \ref{fig:fig1}. If these wavelengths are indeed sampling the wind at its terminal velocity then a reason for this overestimation may be that the wind is cooling closer in than that predicted by the existing model, which assumes a constant temperature of 8,000 K out to $\sim$10 $R_{\star}$. The main mechanism for this cooling would be adiabatic expansion \citep{2011ASPC..448..691O} and would cause lower electron densities than those predicted by the existing model due to larger recombination rates. To investigate this possibility further we adjusted the existing model to include a temperature power-law falloff of the form
\begin{equation}
T_{e}(r)= T_{e}(r_{1})\left(\frac{r_{1}}{r}\right)^{1.65},
\label{eq:eq2}
\end{equation}
at some distance $r_{1}$, from the star. The temperature coefficient is that obtained from our new VLA data assuming a steady flow (see Figure \ref{fig:fig4}). To calculate the new $n_{\rm{e}}$ and $n_{\rm{ion}}$ densities in the wind regime where this temperature falloff occurs, we used the analytical expression of \cite{1986ApJ...306..605G} to calculate the hydrogen ionization fraction, $x_{\rm{HII}}=n_{\rm{HII}}/n_{\rm{H}}$. In doing so we need to make a number of assumptions about the wind properties from $r_{1}$ onwards.
\item 1. Assume a conserved mass outflow i.e., $n_{\rm{H}}(r)=C/r^2$ where $n_{\rm{H}}$ is the total hydrogen number density and $C$ is a constant proportional to the ratio of the mass loss rate divided by the terminal velocity. For $\alpha$ Boo, $C = 1.5 \times 10^{32} $ cm$^{-1}$.
\item 2. All ionization processes cease beyond $r_{1}$(\textbf{GMH}).
\item 3. Consider only radiative recombination of H and include the temperature variation of the recombination coefficient $\alpha _{b}$ which excludes captures to the n=1 level \citep{1978ppim.book.....S}. The recombination coefficient varies with temperature according the relationship
\begin{equation}
\alpha _{b} = \alpha _{b}(r_{1})\left[\frac{T(r_1)}{T(r)}\right]^{0.77},
\label{eq:eq3}
\end{equation}
where the power law coefficient is obtained by finding the slope of the best fit line through the recombination coefficients between 1,000 K and 16,000 K defined in \cite{1978ppim.book.....S}.
\item 4. The heavy ion fraction is constant, $x_{\rm{ion}}=n_{\rm{ion}}/n_{\rm{H}}=10^{-4}$.
\item Using these assumptions it can be shown that the ionization fraction beyond $r_{1}$ is given by \citep{1986ApJ...306..605G}
\begin{equation}
x_{\rm{HII}}(r)= \frac{x_{\rm{HII}}(r_1)x_{\rm{ion}}e^{-I(r)}}{x_{\rm{ion}}+x_{\rm{HII}}(r_1)[1 - e^{-I(r)}]}
\label{eq:eq4}
\end{equation}
where
\begin{equation}
I(r) = 2.38 \times 10^{-3} \left[\left( \frac{r_{1}}{r}\right)^{-0.19} -1 \right], \ \rm{and} \ \  r \geq r_{1}.
\label{eq:eq5}
\end{equation}
\\
By altering the existing atmospheric model so that it now has a narrow temperature  plateau of $T_e = 10,000$ K between 1.2 and 2.3 $R_{\star}$, a temperature profile and a density profile governed by given by Equation \ref{eq:eq2}  and Equation \ref{eq:eq4} beyond $r_{1}$ = 2.3 $R_{\star}$ respectively, then we get  good agreement with our new long wavelength VLA data as shown in Figure \ref{fig:fig1}. This new \textit{hybrid} model, which still has the original ionization fraction of $x_{\rm{H}} \approx 0.5$ inside 2.3 $R_{\star}$, now contains an  initial rapid decrease in $x_{\rm{H}}$ post 2.3 $R_{\star}$ which then tails off to a constant value of $\sim$0.04 beyond $\sim$10$R_{\star}$.

Encouraging as it is that such a simple analytical model can reproduce values close to the observed radio fluxes at long wavelengths, it must be stressed that this \textit{hybrid} model is just a first order approximation. It assumes that the excess slope from the radio spectrum is a result of rapid cooling alone. It still does not reproduce the radio fluxes at wavelengths shorter than $\sim$3 cm and therefore a new atmospheric model is still required that can reproduce all of the observed fluxe densities. To do so, the non-trivial task of simultaneously solving the non-local radiative transfer equation and atomic rate equations which include advection will be required.

\section{CONCLUSIONS}
We have presented the most comprehensive set of multi-wavelength radio continuum observations of two standard luminosity class III red giants to date. This is the first time such classes of stars have been detected longword of 6 cm. Such long wavelength detections are crucial if one wants to study the outer environments of these partially ionized stellar outflows. Our observations were carried out with the VLA during its commissioning phase when only a fraction of the now available bandwidth was at our disposal. The continuous bandwidth coverage between 1 and 50 GHz of the new VLA will allow fast detections of historically weak or undetectable radio continuum luminosity class III red giants at both long and short wavelengths. Previous upper limits will be replaced by firm detections allowing a greater understanding of their atmospheric properties.

>>>>

\subsection{Radio Spectral Indices} \label{disc:disc3}
The radio emission from non-dusty K spectral-type red giants is due to thermal free-free emission in their partially ionized outflows. The radio flux density-frequency relationship for these type of stars is always found to be intermediate between that expected from the stellar disk emission where $\alpha$ follows the Rayleigh-Jeans tail of the Planck function (i.e., $\alpha = +2$) and that from an optically thin plasma ($\alpha = -0.1$). It can be shown that the expected radio spectrum from a spherically symmetric, steady isothermal constant velocity and ionization fraction flow varies as $\nu ^{0.6}$ \citep{1975MNRAS.170...41W,1975AA....39..217O,1975AA....39....1P}. If we relax some of these assumptions about the outflow and instead assume that the electron density and temperature vary as a function of distance from the star $r$, and have the power-law form $n_{e} \propto r^{-p}$ and $T_{e} \propto r^{-n}$ respectively, then

@@

\begin{equation}
\alpha = \frac{4p -6.2 -0.6n}{2p-1-1.35n}
\label{eq:eq1}
\end{equation}
\citep[e.g.,][]{1987ApJ...312..813S}. (always spherical,steady but not ..)

@@

The higher degree of ionization in the mass outflow of $\alpha$ Boo in comparison to $\alpha$ Tau makes a substantial difference to the size of the stars effective radius as can be seen in Figure \ref{fig:fig4}. At 6, 13, and 20 cm the effective radius of $\alpha$ Boo at $\tau _{\rm{rad}}$=1/3 is predicted to be 1.6, 2.8, and 3.7 $R_{\star}$ but is only $\sim$1.1 $R_{\star}$ at 6 and 13 cm for $\alpha$ Tau. \cite{1998ApJ...503..396R} predict that $\alpha$ Tau's wind reaches $\sim$80\% of its terminal velocity by 3 $R_{\star}$ so it is highly unlikely that even our longest wavelength VLA measurements of $\alpha$ Tau sample its outer wind. For $\alpha$ Boo however, \cite{1985pssl.proc..351D} predicts that the wind has reached its terminal velocity by $\sim$2 $R_{\star}$ so based on this model our longest VLA measurements probable sample the wind at, or very close to, its terminal velocity. Assuming this to be the case, Equation \ref{eq:eq1} then indicates that $\alpha$ Boo's steady outflow undergoes rapid cooling at a few stellar radii. The derived temperature coefficient of 1.65 points to a temperature power-law falloff exceeding that from a wind adiabatically cooling with no heat source (i.e. n=1.33 in this case) with the excess cooling being possible due to recombination of H$^{+}$ or heavier ions. Finally, if the wind ionization balance has not yet become \textit{frozen-in} in the region of $\alpha$ Boo's wind where the radio emission emanates from then the excess slope of the spectral index could be due to a combination of both cooling and changing ionization balance. In this scenario the temperature coefficient may be smaller then our derived value here.

@@

A failure of the existing atmospheric model for $\alpha$ Boo is that is overestimates the radio fluxes at long VLA wavelengths which sample its outer atmosphere, as clearly shown in Figure \ref{fig:fig1}. If these wavelengths are indeed sampling the wind at its terminal velocity then a reason for this overestimation may be that the wind is cooling closer in than that predicted by the existing model, which assumes a constant temperature of 8,000 K out to $\sim$10 $R_{\star}$. The main mechanism for this cooling would be adiabatic expansion \citep{2011ASPC..448..691O} and would cause lower electron densities than those predicted by the existing model due to larger recombination rates. To investigate this possibility further we adjusted the existing model to include a temperature power-law falloff of the form
\begin{equation}
T_{e}(r)= T_{e}(r_{1})\left(\frac{r_{1}}{r}\right)^{1.65},
\label{eq:eq2}
\end{equation}
at some distance $r_{1}$, from the star. The temperature coefficient is that obtained from our new VLA data assuming a steady flow (see Figure \ref{fig:fig4}). To calculate the new $n_{\rm{e}}$ and $n_{\rm{ion}}$ densities in the wind regime where this temperature falloff occurs, we used the analytical expression of \cite{1986ApJ...306..605G} to calculate the hydrogen ionization fraction, $x_{\rm{HII}}=n_{\rm{HII}}/n_{\rm{H}}$. In doing so we need to make a number of assumptions about the wind properties from a distnace $r_{1}$ from the star onwards, namely:
\item 1. A conserved mass outflow i.e., $n_{\rm{H}}(r)=C/r^2$ where $n_{\rm{H}}$ is the total hydrogen number density and $C$ is a constant proportional to the ratio of the mass loss rate divided by the terminal velocity. For $\alpha$ Boo, $C = 1.5 \times 10^{32} $ cm$^{-1}$.
\item 2. All ionization processes cease beyond $r_{1}$. The ionization of hydrogen in the chromosphere and wind is a two stage process: the $n = 2$ level is excited by electron collisions and Lyman-alpha scattering, followed by photoionization by the optically thin Balmer continuum. When the temperatures begins to decrease in the wind the collisional excitation rate and thus ionization rate decrease rapidly.
\item 3. Consider only radiative recombination of H and include the temperature variation of the recombination coefficient $\alpha _{b}$ which excludes captures to the n=1 level \citep{1978ppim.book.....S}. The recombination coefficient varies with temperature as
\begin{equation}
\alpha _{b} = \alpha _{b}(r_{1})\left[\frac{T(r_1)}{T(r)}\right]^{0.77},
\label{eq:eq3}
\end{equation}
where the power law coefficient is obtained by finding the slope of the best fit line through the recombination coefficients between 1,000 K and 16,000 K defined in \cite{1978ppim.book.....S}.
\item 4. A fixed ion contribution from from metals with a low first ionization potential, $x_{\rm{ion}}=n_{\rm{ion}}/n_{\rm{H}}=10^{-4}$, as these are easily ionized in the outflow.
\item Using these assumptions it can be shown that the ionization fraction beyond $r_{1}$ is given by \citep{1986ApJ...306..605G}
\begin{equation}
x_{\rm{HII}}(r)= \frac{x_{\rm{HII}}(r_1)x_{\rm{ion}}e^{-I(r)}}{x_{\rm{ion}}+x_{\rm{HII}}(r_1)[1 - e^{-I(r)}]}
\label{eq:eq4}
\end{equation}
where
\begin{equation}
I(r) = 2.38 \times 10^{-3} \left[\left( \frac{r_{1}}{r}\right)^{-0.19} -1 \right], \ \rm{and} \ \  r \geq r_{1}.
\label{eq:eq5}
\end{equation}
\\
By altering the existing atmospheric model so that it now has a narrow temperature  plateau of $T_e = 10,000$ K between 1.2 and 2.3 $R_{\star}$, a temperature profile and a density profile governed by given by Equation \ref{eq:eq2}  and Equation \ref{eq:eq4} beyond $r_{1}$ = 2.3 $R_{\star}$ respectively, then we get  good agreement with our new long wavelength VLA data as shown in Figure \ref{fig:fig1}. This new \textit{hybrid} model, which still has the original ionization fraction of $x_{\rm{H}} \approx 0.5$ inside 2.3 $R_{\star}$, now contains an  initial rapid decrease in $x_{\rm{H}}$ post 2.3 $R_{\star}$ which then tails off to a constant value of $\sim$0.04 beyond $\sim$10 $R_{\star}$.

Encouraging as it is that such a simple analytical model can reproduce values close to the observed radio fluxes at long wavelengths, it must be stressed that this \textit{hybrid} model is just a first order approximation. It assumes that the excess slope from the radio spectrum is a result of rapid cooling alone. It still does not reproduce the radio fluxes at wavelengths shorter than $\sim$3 cm and therefore a new atmospheric model is still required that can reproduce all of the observed fluxe densities. To do so, the non-trivial task of simultaneously solving the radiative transfer equation and non-LTE atomic level populations which include advection will be required.

\section{CONCLUSIONS}
We have presented the most comprehensive set of multi-wavelength radio continuum observations of two standard luminosity class III red giants to date. This is the first time such class of stars have been detected longword of 6 cm. Such long wavelength detections are crucial if one wants to study the outer environments of these partially ionized stellar outflows. Our observations were carried out with the VLA during its commissioning phase when only a fraction of the now available bandwidth was at our disposal. The continuous bandwidth coverage between 1 and 50 GHz of the new VLA will allow fast detections of historically weak or undetectable radio continuum luminosity class III red giants at both long and short wavelengths. Previous upper limits will be replaced by firm detections allowing a greater understanding of their atmospheric properties.
